---
title: "Data Cleaning & Setup"
author: "Lucy Whitmore"
date: "1/25/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


packages <-  c("tidyverse", "tidymodels", "nlme","broom")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())))  
}
lapply(packages, library, character.only = TRUE)


```

# Import data, select for recommended inclusion, join
```{r}
datapath = "/Volumes/devbrainlab/ABCD_Data/ABCD5pt1/core/imaging/"
#directory for saved data is "/Volumes/devbrainlab/

# could filter the inclusion, then join by id & wave (drop other columns from inclusion)
struc_include5<-rio::import(paste0(datapath, "mri_y_qc_incl.csv")) %>%
  select(src_subject_id, eventname, imgincl_t1w_include) %>% 
  #mutate(imgincl_t1w_include = as.numeric(imgincl_t1w_include)) %>% 
  filter(imgincl_t1w_include == 1) 

#qc_info5 <- rio::import("/Volumes/devbrainlab/ABCD_Data/ABCD5pt0/core/imaging/mri_y_qc_man_post_t2w.csv") %>%
#  select(src_subject_id, eventname, imgincl_t1w_include)

#Load structural data
# Volume
smri_vol <- rio::import(paste0(datapath,"mri_y_smr_vol_dsk.csv")) # %>%
 # select(subjectkey, src_subject_id, interview_age, sex, eventname, matches("vol")) 
# note: might just select all, don't need to match for volume anymore  

# Area
smri_area <- rio::import(paste0(datapath,"mri_y_smr_area_dsk.csv"))# %>%
  #select(subjectkey, src_subject_id, interview_age, sex, eventname, matches("area")) 
  
# Thickness
smri_thickness <- rio::import(paste0(datapath,"mri_y_smr_thk_dsk.csv")) #%>%
  #select(subjectkey, src_subject_id, interview_age, sex, eventname, matches("thick"))
 
# Subcortical Volume 
smri_subvol <- rio::import(paste0(datapath,"mri_y_smr_vol_aseg.csv")) %>%
  select(-c(smri_vol_scs_lesionlh, smri_vol_scs_lesionrh))
 # select(subjectkey, src_subject_id, interview_age, sex, eventname, matches("vol"))

# Scanner Info
mri_info5 <- rio::import(paste0(datapath,"mri_y_adm_info.csv")) %>% 
  select(src_subject_id, eventname, mri_info_deviceserialnumber) #, mri_info_manufacturer, mri_info_manufacturersmn)

# Demographic Info (age)
demo <- rio::import("/Volumes/devbrainlab/ABCD_Data/ABCD5pt0/core/abcd-general/abcd_y_lt.csv") %>% 
  select(src_subject_id, eventname, interview_age, rel_family_id) %>% 
  mutate(interview_age = interview_age/12)
```


# Filter for usable data
```{r}
# Cortical Volume
# join cortical and subcortical smri
good_mri_vol <- left_join(struc_include5, smri_vol, by = c("src_subject_id", "eventname"))
good_mri_vol <- left_join(good_mri_vol, mri_info5, by = c("src_subject_id", "eventname")) %>% 
  mutate(src_subject_id = as.factor(src_subject_id),
         eventname = as.factor(eventname), 
         mri_info_deviceserialnumber = as.factor(mri_info_deviceserialnumber)) %>% 
  select(-imgincl_t1w_include)

volume <- left_join(good_mri_vol, demo, c("src_subject_id", "eventname"))
#need to decide whether to bind volume and subcortical volume before combat

# Subcortical Volume
good_mri_subvol <- left_join(struc_include5, smri_subvol, by = c("src_subject_id", "eventname"))
good_mri_subvol <- left_join(good_mri_subvol, mri_info5, by = c("src_subject_id", "eventname")) %>% 
  mutate(src_subject_id = as.factor(src_subject_id),
         eventname = as.factor(eventname), 
         mri_info_deviceserialnumber = as.factor(mri_info_deviceserialnumber)) %>% 
  select(-imgincl_t1w_include)

subvolume <- left_join(good_mri_subvol, demo, c("src_subject_id", "eventname")) %>% 
  select(-c(smri_vol_scs_wmhintlh, smri_vol_scs_wmhintrh))


# Cortical Area
good_mri_area <- left_join(struc_include5, smri_area, by = c("src_subject_id", "eventname"))
good_mri_area <- left_join(good_mri_area, mri_info5, by = c("src_subject_id", "eventname"))%>% 
  mutate(src_subject_id = as.factor(src_subject_id),
         eventname = as.factor(eventname), 
         mri_info_deviceserialnumber = as.factor(mri_info_deviceserialnumber)) %>% 
  select(-imgincl_t1w_include)

area <- left_join(good_mri_area, demo, c("src_subject_id", "eventname"))
```

Join all measures and save
```{r}
# merge back into one dataframe that can be passed into model training
all_mri <- left_join(area, volume, by = c('src_subject_id', 'eventname','mri_info_deviceserialnumber', 'rel_family_id', 'interview_age'))
all_mri <- left_join(all_mri, subvolume, by = c('src_subject_id', 'eventname','mri_info_deviceserialnumber', 'rel_family_id', 'interview_age')) %>% 
  relocate(c('interview_age', 'rel_family_id', 'mri_info_deviceserialnumber'), .before=smri_area_cdk_banksstslh)

save(all_mri, file="all_mri.Rda")

#rm(list=setdiff(ls(), "all_mri"))

```



## Import Data
```{r}

all_mri <- rio::import("/Volumes/devbrainlab/Lucy/BrainAGE/ChangeScoresCapstone/all_mri.Rda")

```

# Cleaning/Prep
```{r}
# split by wave
# probably keep subject id at least
# (join by that later)
mri_baseline <- all_mri %>% 
  filter(eventname == "baseline_year_1_arm_1")  %>% 
  select(-c(mri_info_deviceserialnumber)) #src_subject_id, eventname, 

#head(mri_baseline %>% select(c(interview_age, interview_age.x, interview_age.y)))

mri_followup_y2 <- all_mri %>% 
  filter(eventname == "2_year_follow_up_y_arm_1")  %>% 
  select(-c(mri_info_deviceserialnumber, rel_family_id))  #src_subject_id, eventname, 

mri_followup_y4 <- all_mri %>% 
  filter(eventname == "4_year_follow_up_y_arm_1")  %>% 
  select(-c(mri_info_deviceserialnumber, rel_family_id)) #src_subject_id, eventname, 


```

## Calculate difference scores

Annualized change approach
```{r}
# Join data
smri_difference_baseline_y2 <- left_join(mri_baseline, mri_followup_y2, by =c("src_subject_id"))

smri_difference_y2_y4 <- left_join(mri_followup_y2, mri_followup_y4, by =c("src_subject_id"))

# Calculate Change
smri_difference_baseline_y2_calc <- smri_difference_baseline_y2 %>%
 select(-c(eventname.x,eventname.y, rel_family_id)) %>% 
 mutate(across(ends_with(".y"), .names = "{col}_diff") - across(ends_with(".x"))) %>%
# rename_with(~ sub("_\\d+", "", .), ends_with("_diff"))# %>%  #removes numeric suffixes 
  rename_with(~ sub("\\.y_diff$", "_diff", .), ends_with(".y_diff")) %>% 
  select(-contains(c('.x', '.y')))

smri_difference_y2_y4_calc <- smri_difference_y2_y4 %>%
 select(-c(eventname.x,eventname.y)) %>% 
 mutate(across(ends_with(".y"), .names = "{col}_diff") - across(ends_with(".x"))) %>%
 rename_with(~ sub("\\.y_diff$", "_diff", .), ends_with(".y_diff")) %>% 
 select(-contains(c('.x', '.y')))


#divide all brain columns by change in interview age
# this gets us per year change (not a %, but numeric change per year)

# Baseline - Year 2
smri_difference_baseline_y2_scaled <- smri_difference_baseline_y2_calc # for some reason have to duplicate or can't have first two columns
smri_difference_baseline_y2_scaled[, -(1:2)] <- sweep(smri_difference_baseline_y2_scaled[, -(1:2)], 1, smri_difference_baseline_y2_scaled [, 2], "/")

smri_difference_baseline_y2_scaled <- smri_difference_baseline_y2_scaled %>% 
  filter(!is.na(interview_age_diff))

# at some point, want to reattach outcome age, but need to decide what that should be, probably T2 age
smri_difference_baseline_y2_scaled <- left_join(smri_difference_baseline_y2_scaled, mri_followup_y2 %>% select(c("src_subject_id", "interview_age")), by="src_subject_id")

smri_difference_baseline_y2_scaled <- left_join(smri_difference_baseline_y2_scaled, mri_baseline %>% select(c("src_subject_id", "rel_family_id")), by="src_subject_id") %>% 
  relocate(c('interview_age', 'rel_family_id'), .before=interview_age_diff)


# Year 2 - Year 4
smri_difference_y2_y4_scaled <- smri_difference_y2_y4_calc # for some reason have to duplicate or can't have first two columns
smri_difference_y2_y4_scaled[, -(1:2)] <- sweep(smri_difference_y2_y4_scaled[, -(1:2)], 1, smri_difference_y2_y4_scaled[, 2], "/")

smri_difference_y2_y4_scaled <- smri_difference_y2_y4_scaled %>% 
  filter(!is.na(interview_age_diff))


# attach family ID and outcome age
smri_difference_y2_y4_scaled <- left_join(smri_difference_y2_y4_scaled, mri_followup_y4 %>% select(c("src_subject_id", "interview_age")), by="src_subject_id")

smri_difference_y2_y4_scaled <- left_join(smri_difference_y2_y4_scaled, mri_baseline %>% select(c("src_subject_id", "rel_family_id")), by="src_subject_id") %>% 
  relocate(c('interview_age', 'rel_family_id'), .before=interview_age_diff)




save(smri_difference_baseline_y2_scaled, file="smri_difference_baseline_y2_scaled.Rda")
save(smri_difference_y2_y4_scaled, file="smri_difference_y2_y4_scaled.Rda")
```


Random slope approach
```{r}
# run models in a loop, extract a random slope for each brain region
# limitation: have to have all 3 waves

#https://github.com/abcdworkshop/lmmtutorialtrack1/blob/main/track1mlmtutorial.Rmd

# make subject id a factor
all_mri <- all_mri %>% 
  mutate(src_subject_id = as.factor(src_subject_id))

# limit to only those with data at all 3 waves
slope_count <- all_mri %>% count(src_subject_id) %>% filter(n >= 3)

all_mri_filtered <- all_mri %>% 
  filter(src_subject_id %in% slope_count$src_subject_id)


## Run lme in a for loop
ranef_df <- data.frame(src_subject_id = unique(all_mri_filtered$src_subject_id))


# Loop over each column in the dataframe (excluding the first 5 columns)
  for (col in names(all_mri_filtered)[6:length(all_mri_filtered)]) {
#for (col in names(all_mri_filtered)[6:25]) {
    print(col)
   # exampleFormula = paste0(col, fixed_part)
  # Fit the linear mixed-effects model for the current column
  fm = lme(as.formula(paste0(col, " ~ interview_age")), method='ML', control =list(msMaxIter = 1000, msMaxEval = 1000, sing.tol=1e-20), random = ~1+interview_age|src_subject_id, data=all_mri_filtered)
  
  # Extract the random effects for each subject
  ranef_values <- ranef(fm)[2]  # Extracting the second column from ranef
  
  # Add the random effects to the dataframe
  ranef_df[[col]] <- ranef_values
}



# convert to regular dataframe
ranef_df_numeric <- as.data.frame(sapply(ranef_df[2:length(ranef_df)] , unlist ))

# add IDs
ranef_df_numeric$src_subject_id = slope_count$src_subject_id 

ranef_df_numeric  <- ranef_df_numeric  %>%
  select(src_subject_id, everything())

#renumber rows
row.names(ranef_df_numeric ) <- 1:nrow(ranef_df_numeric)

# add interview age, family ID
slopes_df <- left_join(ranef_df_numeric, all_mri %>% filter(eventname == "4_year_follow_up_y_arm_1") %>% select(c("src_subject_id", "interview_age")), by="src_subject_id")

slopes_df <- left_join(slopes_df, all_mri %>% filter(eventname == "baseline_year_1_arm_1") %>% select(c("src_subject_id", "rel_family_id")), by="src_subject_id")

slopes_df  <- slopes_df  %>%
  select(src_subject_id, interview_age, rel_family_id, everything())


save(slopes_df, file="slopes_df.Rda")

```

check for sample size of annualized change vs random slope approach
```{r}
# Subtraction method
nrow(smri_difference_baseline_y2_scaled)
nrow(smri_difference_y2_y4_scaled)

# Change score
nrow(ranef_df)
```

## Limit to same participants in both data sets & join both waves?
```{r}
smri_difference_baseline_y2_scaled  <- rio::import("/Volumes/devbrainlab/Lucy/BrainAGE/ChangeScoresCapstone/smri_difference_baseline_y2_scaled.Rda")

smri_difference_y2_y4_scaled <- rio::import("/Volumes/devbrainlab/Lucy/BrainAGE/ChangeScoresCapstone/smri_difference_y2_y4_scaled.Rda")

# filter annualized change for IDs in random slope
smri_difference_baseline_y2_scaled  <- smri_difference_baseline_y2_scaled %>% 
  filter(src_subject_id %in% slopes_df$src_subject_id)


smri_difference_y2_y4_scaled <- smri_difference_y2_y4_scaled %>% 
  filter(src_subject_id %in% slopes_df$src_subject_id)


# Join both waves together
# probably mutate names first to tell apart
colnames(smri_difference_baseline_y2_scaled)[c(2, 4:length(smri_difference_baseline_y2_scaled))] <- paste(colnames(smri_difference_baseline_y2_scaled)[c(2, 4:length(smri_difference_baseline_y2_scaled))], "y2", sep = "_")

colnames(smri_difference_y2_y4_scaled)[c(2, 4:length(smri_difference_y2_y4_scaled))] <- paste(colnames(smri_difference_y2_y4_scaled)[c(2, 4:length(smri_difference_y2_y4_scaled))], "y4", sep = "_")


annualized_change <- left_join(smri_difference_baseline_y2_scaled %>% select(-c("interview_age_diff_y2", "interview_age_y2")), smri_difference_y2_y4_scaled  %>% select(-c("interview_age_diff_y4")), by = c("src_subject_id", "rel_family_id")) %>% 
  relocate(interview_age_y4, .before = smri_area_cdk_banksstslh_diff_y2)

save(annualized_change, file="annualized_change.Rda")

```

## Split for train/test (slopes)
```{r}
# set seed
# make sure siblings are put into same split
# need to import family info

# check to see what max number of people per group is
slopes_df %>% count(rel_family_id)
# probably graph distribution

set.seed(42)

slopes_split <- group_initial_split(
  slopes_df,
  group = rel_family_id,
  prop = 0.80 #,
  # matching age distributions across train and test set
 # strata = "interview_age"
)

slopes_train <- training(slopes_split)
slopes_test <- testing(slopes_split)

# graph age distribution of both splits
hist(slopes_train$interview_age, breaks = 5)
hist(slopes_test$interview_age, breaks = 5)

t.test(slopes_train$interview_age, slopes_test$interview_age)



save(slopes_train, file="slopes_train.Rda")
save(slopes_test, file="slopes_test.Rda")

```

## Split for train/test (annualized change)
```{r}

annualized_train <- annualized_change %>% 
  filter(src_subject_id %in% slopes_train$src_subject_id)
  
annualized_test <- annualized_change %>% 
  filter(src_subject_id %in% slopes_test$src_subject_id)

save(annualized_train, file="annualized_train.Rda")
save(annualized_test, file="annualized_test.Rda")
```

